Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean throughout the United States, and from Mexico, France, and Australia

### 2017-06-29 doc/manuscript/manuscript.Rmd
update publication

All data processing and analyses were performed in R version 3.4.0 [@R].
Of the 16 microsatellite loci, five included compound repeats, making allele scoring difficult. 
These loci were removed for the reported statistics, but were processed and included in data analyses to ensure the integrity of the results with 11 loci as compared to 16.
To control for non-independent samples (i.e. those sampled from the same location, region, year, and plant), data were clone-corrected for subsequent analyses.
The clone-corrected data set included 318 isolates.
All analyses were performed on this data set unless otherwise mentioned.
All data and analyses are openly available and reproducible at https://github.com/everhartlab/sclerotinia_366/

@@

<!--  ZNK: I think figures 1-4 are a bit redundant with the tables in the text. It might be better to simply remove them. -->
A total of 366 isolates were collected from 2003 to 2012 from diseased dry bean plants in 11 states in the US, Australia (2 mainland; 4 Tasmania), Mexico, and France **(Fig. 1)**.
Most isolates came from MI (62), ND (60), WA (59), NE (47), and CO (42).  Other regions in lesser representation were France (24), MX (18), OR (17), CA (16), MN (11), Australia (6), WI (2), ID (1), and NY (1).
Four of the Australian isolates were collected from Tasmania while the other two came from Victoria state.
There were no samples collected in 2006 and 2011.
The highest number of samples were collected in 2005 (N = 63) and lowest in 2012 (N = 4).
Pathogen isolates represented 26 commercial cultivars and breeding lines with some representing unknown dry bean hosts **(Fig. 2)**.
We investigated the population association with host for cultivars where more than 10 isolates were obtained, which included G122, Beryl, Bunsi, Yellow, Pinto, and Merlot.
These six cultivars had a total of 238 *S. sclerotiorum* isolates (of 366 total isolates collected) belonging to 130 MLGs.
Of these MLGs, 53 were singletons, while the rest were obtained from more than one host cultivar; only one private allele on locus 55-4 was identified, belonging to the *S. sclerotiorum* population from the cultivar 'Bunsi'.
The number of unique MLGs recorded from each host is shown in Table 3.
The majority of MLGs reported were found on multiple host cultivars and only a few unique to a cultivar.
For instance, there were only 15 unique MLGs within the population of 68 *S. sclerotiorum* isolates obtained from the 'Beryl' cultivar.

There were a total of 165 multilocus genotypes in the isolate collection.
The most abundant MLG represented 27 isolates (Fig. 3 and Table 4) across five regions (NE, MI, WA, CO, and ND).
Twenty isolates of this MLG were identified from ND.
The genotypes that had either highest geographical distribution or found in high frequency are given in Table 4.
MLG 163 was identified in NE, MI, CO and ND, while MLG 165 was identified in MN and all states aforementioned, except for ND.
Isolates with MLG 15 were identified from populations in MI, WA, CO, and ND.
The MLGs 163, 165 and 15 had 15, 7, and 4 individuals across these populations, respectively.
The highest number of clones was recorded from ND, NE, and MI, which included 20, 10 and 7 isolates, respectively (Table 4).

@@

The number of observed alleles per locus ranged anywhere from 2 to 10 (Table 2).
All loci were informative at cut off level of 2/N where N was the total number of individuals.
The genetic diversity indices of *S. sclerotiorum* populations from different states of the USA, France, Mexico and Australia based on SSR analysis are given in Table 5.
The 165 unique MLGs identified among all 366 isolates are given in Supplementary Table 1.
Gene diversity ($H_{exp}$) of populations from different regions ranged from 0.259 to 0.596 (Table 5).
Evenness of genotypes of all regions was high, with the exception of ND, where a single genotype was represented by 20 isolates.
There was one private allele in the WA population (204 bp allele of 55-4 locus) and two private alleles in the ND population (two 366 bp alleles in 9-2 locus and two 220 bp alleles in 12-2 locus).

@@

We tested the clone corrected data for evidence of sexual outcrossing within *S. sclerotiorum* populations in different regions and different years/region.
CA showed evidence of sexual outcrossing (Ia = 0.08, P.Ia = 0.317, $\bar{r}_d$ = 0.008, P.$\bar{r}_d$ = 0.317) while all other regions implied clonal propagation (data not shown).
Five populations showed evidence of sexual outcrossing: CA in 2004, FR in 2004 and 2005, OR in 2003, and WA in 2008.
However, after clone-correction these populations had 6-7 isolates/population (Table 7) and therefore, may not be sufficient to make inferences about reproductive mode.
Only exception was the WA_2008 population, which consisted of 12 isolates.
Within sampling locations white mold nurseries (wmn) were tested for linkage disequilibrium since same wmn fields were used for sampling over different years.
Only CA wmn showed evidence of sexual reproduction ( = 0.008; P = 0.29) while all other wmn locations suggested clonal reproduction.
The DAPC analyses of WA, and MI isolates from same wmn fields over different years are in Fig. 10.
Since *S. sclerotiorum* isolates are predominantly clonal, we wanted to explore whether there was any introduction of new isolates from other fields or if there were evidence of sexual reproduction.
The AMOVA results for different clusters of Fig. 10A revealed significance difference between WA 2008 isolates from the rest ($\Phi_{PT}$ $\geq$ 0.206, P = 0.001).
The 2003 and 2005 isolates were genetically similar ($\Phi_{PT}$ = 0.056, P = 0.129).
MI isolates from 2008 and 2003 (Fig. 10B) showed significant difference ($\Phi_{PT}$ = 0.127, P = 0.04) while 2005 isolates did not differ molecularly from both 2008 and 2003 isolates ($\Phi_{PT}$ $\leq$ 0.078, P $\geq$ 0.076).
NE wmn isolates sampled in 2004 and 2008 were genetically similar ($\Phi_{PT}$ = 0.042, P = 0.179) (DAPC results not shown).
The standardized index of association of wmn isolates from each region (i.e.
MI, WA, and NE) for above mentioned sampling years suggested clonal reproduction ( $\geq$ 0.05, P $\leq$ 0.002).

>>>>

All data processing and analyses were performed in R version 3.4.0 [@R].
Of the 16 microsatellite loci, five included compound repeats, making allele scoring difficult. 
These loci were removed for the reported statistics, but were processed and included in data analyses to ensure the integrity of the results with 11 loci as compared to 16.
To control for non-independent samples (i.e. those sampled from the same location, region, year, and plant), data were clone-corrected for subsequent analyses.
The clone-corrected data set included 318 isolates.
All analyses were performed on this data set unless otherwise mentioned.
All data and analyses are openly available and reproducible at https://github.com/everhartlab/sclerotinia-366/

@@

![
Barplot showing abundance of each Mycelial Compatibility Group (MCG) in descending order.
White bars represent sample counts and grey bars represent counts of unique multilocus genotypes (MLG).
The transparency associated with the grey bars represent the evenness of the distribution of the MLGs within a given MCG as shown in the inset histograms.
](../../results/figures/publication/Figure1Z.pdf){#barplot width=50%}

<!--  ZNK: I think figures 1-4 are a bit redundant with the tables in the text. It might be better to simply remove them. -->
A total of 366 isolates were collected from 2003 to 2012 from diseased dry bean plants in 11 states in the US, Australia (2 mainland; 4 Tasmania), Mexico, and France **(Fig. 1)**.
Most isolates came from MI (62), ND (60), WA (59), NE (47), and CO (42).  Other regions in lesser representation were France (24), MX (18), OR (17), CA (16), MN (11), Australia (6), WI (2), ID (1), and NY (1).
Four of the Australian isolates were collected from Tasmania while the other two came from Victoria state.
There were no samples collected in 2006 and 2011.
The highest number of samples were collected in 2005 (N = 63) and lowest in 2012 (N = 4).

```{r table3, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/mlg_table.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character())) %>%
  rename(`$P_{sex}$` = Psex) %>% 
  knitr::kable(format = "pandoc")
```

Table: (\#tab:mlg-table) The 5 most abundant Multilocus Genotypes (MLG) with
the probability of second encounter ($P_{sex}$), Mycelial Compatibility Groups (MCG), and Regions with sample sizes in parentheses.

Pathogen isolates represented 26 commercial cultivars and breeding lines with some representing unknown dry bean hosts **(Fig. 2)**.
We investigated the population association with host for cultivars where more than 10 isolates were obtained, which included G122, Beryl, Bunsi, Yellow, Pinto, and Merlot.
These six cultivars had a total of 238 *S. sclerotiorum* isolates (of 366 total isolates collected) belonging to 130 MLGs.
Of these MLGs, 53 were singletons, while the rest were obtained from more than one host cultivar; only one private allele on locus 55-4 was identified, belonging to the *S. sclerotiorum* population from the cultivar 'Bunsi'.
The five most abundant MLGs with associated MCGs are shown in Table \@ref(tab:mlg-table).
The majority of MLGs reported were found on multiple host cultivars and only a few unique to a cultivar.
For instance, there were only 15 unique MLGs within the population of 68 *S. sclerotiorum* isolates obtained from the 'Beryl' cultivar.

There were a total of 165 multilocus genotypes in the isolate collection.
The most abundant MLG represented 27 isolates (Fig. \@ref(barplot) and Table 4) across five regions (NE, MI, WA, CO, and ND).
Twenty isolates of this MLG were identified from ND.
The genotypes that had either highest geographical distribution or found in high frequency are given in Table 4.
MLG 163 was identified in NE, MI, CO and ND, while MLG 165 was identified in MN and all states aforementioned, except for ND.
Isolates with MLG 15 were identified from populations in MI, WA, CO, and ND.
The MLGs 163, 165 and 15 had 15, 7, and 4 individuals across these populations, respectively.
The highest number of clones was recorded from ND, NE, and MI, which included 20, 10 and 7 isolates, respectively **(Table 4)**.

@@

----------------------------------------------
Locus  Range    No. alleles $H_{exp}$ Evenness 
------ ------- ------------ --------- --------
5-2    318-324            4      0.45     0.62

6-2    483-495            3      0.64     0.95

7-2    158-174            7      0.73     0.76

8-3    244-270            7      0.74     0.79

9-2    360-382            9      0.35     0.41

12-2   214-222            5      0.58     0.78

17-3   342-363            7      0.55     0.53

20-3   280-282            2      0.05     0.42

55-4   153-216           10      0.72     0.66

110-4  370-386            5      0.76     0.91

114-4  339-416           10      0.83     0.80

*mean* 153-495         6.27      0.58     0.69

----------------------------------------------

Table: (\#tab:locus-stats) Allelic diversity on full data set at loci used in this study. $H_{exp}$ = Nei's Gene Diversity [@nei1978estimation].

The number of observed alleles per locus ranged anywhere from 2 to 10 (Table \@ref(tab:locus-stats)).
All loci were informative at cut off level of 2/N where N was the total number of individuals.
The genetic diversity indices of *S. sclerotiorum* populations from different states of the USA, France, Mexico and Australia based on SSR analysis are given in Table 5.
The 165 unique MLGs identified among all 366 isolates are given in Supplementary Table 1.
Gene diversity ($H_{exp}$) of populations from different regions ranged from 0.259 to 0.596 **(Table 5)**.
Evenness of genotypes of all regions was high, with the exception of ND, where a single genotype was represented by 20 isolates.
There was one private allele in the WA population (204 bp allele of 55-4 locus) and two private alleles in the ND population (two 366 bp alleles in 9-2 locus and two 220 bp alleles in 12-2 locus).

@@

We tested the clone corrected data for evidence of sexual outcrossing within *S. sclerotiorum* populations in different regions and different years/region.
CA showed evidence of sexual outcrossing (Ia = 0.08, P.Ia = 0.317, $\bar{r}_d$ = 0.008, P.$\bar{r}_d$ = 0.317) while all other regions implied clonal propagation (data not shown).
Five populations showed evidence of sexual outcrossing: CA in 2004, FR in 2004 and 2005, OR in 2003, and WA in 2008.
However, after clone-correction these populations had 6-7 isolates/population **(Table 7)** and therefore, may not be sufficient to make inferences about reproductive mode.
Only exception was the WA_2008 population, which consisted of 12 isolates.
Within sampling locations white mold nurseries (wmn) were tested for linkage disequilibrium since same wmn fields were used for sampling over different years.
Only CA wmn showed evidence of sexual reproduction ( = 0.008; P = 0.29) while all other wmn locations suggested clonal reproduction.
The DAPC analyses of WA, and MI isolates from same wmn fields over different years are in Fig. 10.
Since *S. sclerotiorum* isolates are predominantly clonal, we wanted to explore whether there was any introduction of new isolates from other fields or if there were evidence of sexual reproduction.
The AMOVA results for different clusters of Fig. 10A revealed significance difference between WA 2008 isolates from the rest ($\Phi_{PT}$ $\geq$ 0.206, P = 0.001).
The 2003 and 2005 isolates were genetically similar ($\Phi_{PT}$ = 0.056, P = 0.129).
MI isolates from 2008 and 2003 (Fig. 10B) showed significant difference ($\Phi_{PT}$ = 0.127, P = 0.04) while 2005 isolates did not differ molecularly from both 2008 and 2003 isolates ($\Phi_{PT}$ $\leq$ 0.078, P $\geq$ 0.076).
NE wmn isolates sampled in 2004 and 2008 were genetically similar ($\Phi_{PT}$ = 0.042, P = 0.179) (DAPC results not shown).
The standardized index of association of wmn isolates from each region (i.e.
MI, WA, and NE) for above mentioned sampling years suggested clonal reproduction ( $\geq$ 0.05, P $\leq$ 0.002).
